Interisland movements of woodland caribou in Newfoundland

### 2020-02-21 paper/manuscript.Rmd
refs refszzzzz

The Fogo Island archipelago, off the coast of Newfoundland, Canada (49°N, 54°W),
is home to approximately 300 woodland caribou. During routine fieldwork on May
30, 2017, we observed an unmarked adult male caribou swim between Perry and
Eastern Indian Islands (Figure 1), a swim of at least 470 m which took ~9
minutes (~52 m per minute). This observation prompted us to investigate whether
swimming behaviour was common among GPS radio-collared caribou in the population
(n = 29 adult females: 1240g, GPS 4400M; Lotek Wireless Inc., Newmarket,
Ontario, Canada). We identified swimming events when two consecutive GPS (2-hour
relocation rates) locations from an individual occurred on different islands.
This region typically experiences pack ice during the winter, so caribou may be
able to travel between islands by walking over the ice. We restricted our GPS
data to the ice-free period of the year (April 1 - November 25) and we
identified three adult female caribou that collectively engaged in 86 swimming
events (Figure 2).

@@

Caribou may swim between islands in the ocean for similar reasons that drive
caribou swimming in freshwater (Leblond et al. 2016). We present two
explanations for these swimming events, forage limitation (Miller 2002) and
predator avoidance (Jeffery et al. 2007), and discuss why these swimming events
appear to be relatively commonplace in the population.

## Forage limitation
Forage limitation and over-grazing is a major concern for caribou populations
(Schaefer et al. 2016; Zamin et al. 2017) and could explain movement between
islands. Conspecific competition for resources driven by reduced foraging
opportunities can decrease reproductive success of adult females (Schaefer et
al. 2016). Forage scarcity has been proposed as a potential reason for caribou
moving between arctic islands on the sea-ice in winter (Miller et al. 1977).
For very small islands, over-grazing by newly arrived caribou may rapidly
deplete forage (Bergerud et al. 1990). Without sea ice, movement from the
mainland to an island, or between islands, requires caribou to assess the
trade-off associated with swimming.

This trade-off can be understood as density-dependent habitat selection governed
by the Ideal Free Distribution (Morris 1987; Bradbury et al. 2015). Ideal Free
Distribution theory predicts a fitness equilibrium: when the density in a given
habitat patch has exceeded the optimum for fitness within that patch, animals
should relocate and settle new habitat patches, so that fitness is equal across
all patches (Bradbury et al. 2015). Islands act as discrete habitat patches in
this case, but the costs associated with swimming create a trade-off when moving
between patches. For caribou in the Fogo Island archipelago, the role of forage
limitation in animal habitat selection patterns and movement remains unknown,
but swimming between islands may in part be explained by forage limitation and
conspecific competition.

>>>>

The Fogo Island archipelago, off the coast of Newfoundland, Canada,
is home to approximately 300 woodland caribou. During routine fieldwork on May
30, 2017, we observed an unmarked adult male caribou swim between Wester and
Eastern Indian Islands (Figure 1), a swim of at least 470 m which took ~9
minutes (~52 m per minute). This observation prompted us to investigate whether
swimming behaviour was common among GPS radio-collared caribou in the population
(n = 29 adult females: 1240g, GPS 4400M; Lotek Wireless Inc., Newmarket,
Ontario, Canada). We identified swimming events when two consecutive GPS (2-hour
relocation rates) locations from an individual occurred on different islands.
This region typically experiences pack ice during the winter, so caribou may be
able to travel between islands by walking over the ice. We restricted our GPS
data to the ice-free period of the year (April 1 - November 25) and we
identified three adult female caribou that collectively engaged in 86 swimming
events (Figure 2).

@@

Caribou may swim between islands in the ocean for similar reasons that drive
caribou swimming in freshwater [@Leblond_2016]. We present two
explanations for these swimming events, forage limitation (Miller 2002) and
predator avoidance (Jeffery et al. 2007), and discuss why these swimming events
appear to be relatively commonplace in the population.

## Forage limitation
Forage limitation and over-grazing is a major concern for caribou populations
[@Schaefer_2016] and could explain movement between
islands. Conspecific competition for resources driven by reduced foraging
opportunities can decrease reproductive success of adult females [@Schaefer_2016]. Forage scarcity has been proposed as a potential reason for caribou
moving between arctic islands on the sea-ice in winter [@Miller_1977].
For very small islands, over-grazing by newly arrived caribou may rapidly
deplete forage [@Bergerud_1990]. Without sea ice, movement from the
mainland to an island, or between islands, requires caribou to assess the
trade-off associated with swimming.

This trade-off can be understood as density-dependent habitat selection governed
by the Ideal Free Distribution [@Morris_1987; @Bradbury_2015]. Ideal Free
Distribution theory predicts a fitness equilibrium: when the density in a given
habitat patch has exceeded the optimum for fitness within that patch, animals
should relocate and settle new habitat patches, so that fitness is equal across
all patches (Bradbury et al. 2015). Islands act as discrete habitat patches in
this case, but the costs associated with swimming create a trade-off when moving
between patches. For caribou in the Fogo Island archipelago, the role of forage
limitation in animal habitat selection patterns and movement remains unknown,
but swimming between islands may in part be explained by forage limitation and
conspecific competition.
